---
title: __Homicides in Mexico (2019 - 2020)__
subtitle: __Exploratory analysis__

author: 
- Fernando Alarid-Escudero, Ph.D., Yadira Peralta, Ph.D.
- Regina Isabel Medina Rosales, Mariana Fernández

output: pdf_document
---

\tableofcontents
\newpage

<!---  Configuración inicial para programación en R --> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_df, include=FALSE}
# load("../../data/df_national.Rdata")
# load("../../data/df_state.Rdata")
# load("../../data/df_gender.Rdata")
# load("../../data/df_age_groups.Rdata")
# load("../../data/df_gender_age.Rdata")
# load("../../data/df_mort_homicides_national_age_group_sex.Rdata")

```


## The challenge of obtaining recent homicide data
Publishing official homicide data entails a rigurous and extense process. Before being accounted as a homicide by INEGI, every single case has to be confirmed by the local authorities as an actual crime. Due to this reason, official homicide data takes long to be open to the public. Normally, official data for a specific year is published one year after it ended. That is the reason why, so far, INEGI has only published homicide data for 1990 to 2018. 

What to do, then, if one wants to study the phenomenon of homice in Mexico in the recent 18 months? Luckyly, the Secretariat of Security and Citizen Protection (SSPC as its acronym in spanish) makes premiliminary homicide data available on its [website](http://www.informeseguridad.cns.gob.mx/). It uses two main sources of information: a daily report produced by an interdisciplinary team (constituted by the own SSCP, the military, the navy and the General Attoney's office); and open sources, constituted by homicide reports in local newspapers. The SSCP publishes daily reports for the numbers obtained from these two sources.

The problem with the reports that the SSCP uploads on its website is that they come in a pdf format, which means that a deep cleaning process is needed. In the first place, the documents have to be webscrapped from the web, then, they have to be transformed into an excel format and checked by a human. The cleaning process also requires ortographic correction and minor but numerous changes. After homicide data is cleaned, population data is added at both the state and municipal level for the estimation of the homicide rate. 

## Source 1: Interistitutional group report 
The daily reports produced by the interinstitutional group consist in a single table that shows the number of killings in each of the 32 Mexican states, as well as the national toll. This is the descriptive analysis produced after data cleaning. 

## Source 2: Open sources









